---
title: "Reproducible research concepts and tools, 2017"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: ioslides_presentation
runtime: shiny
---

```{r setup,echo=FALSE}
suppressPackageStartupMessages({
library(png)
library(grid)
})
```

## The three themes underlying reproducibility research

1) providing **code** and **data** and *environment* to **independent parties**
to **diminish risk** of analyses that are **not reproducible**

2) fortifying **criteria of statistical soundness** of analyses (study interpretations) to **control risk** of **non-replicability** of **studies**

3) doing 1) and 2) in ways that are cost-effective

## Road map of talk

* Basic terminology
* Reflections on GWAS replication and genetic medicine
* Replication concepts in the NAS workshop
* Reproducible research recommendations of ASA
* Some case studies
* Some github exercises


## Basic definitions

1) Analysis = software + data + environment + invocations

2) Reproducible analysis = analysis that can be carried out by independent parties

3) Extensible analysis = analysis supporting independent variations

4) Study = Design + implementation + analysis

5) Replicable study = A study that, when executed by independent parties,
produces statistically compatible interpretations

## Open questions

1) What is an environment for an analysis?

2) What are independent parties? (See Kahneman's [concept of "replication etiquette"](https://www.scribd.com/document/225285909/Kahneman-Commentary))

3) What are independent variations on an analysis and why are they important?

4) When are two interpretations of related or identical studies statistically compatible?  Reformulate in terms of the standards of replicability for a given field.  Example: GWAS catalog concept of replicated finding (see, e.g., P. Kraft et al., Stat. Sci. 2009).

## GWAS: [p-values are not created equal](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2865141/)

```{r foo,echo=FALSE}
im = readPNG("images/KraftGWASreplication.png")
grid.raster(im,height=1.2)
```

## GWAS replication recommendations in Kraft et al. 2009

* Test the same marker 
* Use the same analytic methods; report
the final model in as much detail as possible so that other investigators
can judge the fit of that model in other datasets
* Try to use the same phenotype

* Issues: 
    * measuring heterogeneity of effects
    * choosing models and studies for metaanalysis
    * understanding failure to replicate

## STREGA: genetic association reporting guidelines

```{r foo2,echo=FALSE}
im = readPNG("images/StregaTitle.png")
grid.raster(im)
```

## Guideline 12
```{r foo3,echo=FALSE}
im = readPNG("images/strega12.png")
grid.raster(im)
```

## Guidelines 13-16
```{r foo4,echo=FALSE}
im = readPNG("images/strega13_16.png")
grid.raster(im)
```

## Upshots

- Replicability is a live concept for population genetics research
- Contributions to literature are managed through guidelines
- The GWAS catalog
